Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-09 ICML-submission/appendix-multiclass-perceptron.tex
Fix formatting in multi-class perceptron appendix

\begin{proof}
Let $M = \sum_{t=1}^T \indicator{\widehat y_t \neq y_t}$ be the number of
mistakes the algorithm makes. Since the $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots,
w_K^{(t)})$ changes only if a mistake is made, we can upper bound $\sum_{i=1}^K
\norm{w_i^{(t)}}^2$ in terms of number of mistakes.
If a mistake happens in round $t$ then
\begin{align*}
\sum_{i=1}^K \norm{w_i^{(t+1)}}^2
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) + \norm{w_{y_t}^{(t)} + x_t}^2 + \norm{w_{\widehat y_t}^{(t)} - x_t}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) + \norm{w_{y_t}^{(t)}}^2 + \norm{w_{\widehat y_t}^{(t)}}^2 + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& = \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 R^2 \; .
\end{align*}
So each time a mistake happens, $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ increases by at most $2R^2$. Thus,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors satisfying the
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We lower bound $\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$. This quantity changes
only when a mistakes happens. If mistake happens in round $t$, we have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(t+1)}}
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{y_t}^*}{w_{y_t}^{(t)} + x_t} + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} - x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{y_t}^* - w_{\widehat y_t}^*}{x_t} \\
& \ge  \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma \; .
\end{align*}
Thus, after $M$ mistakes,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma M \; .
$$
We upper bound the left hand side by using Cauchy-Schwartz inequality twice and
the condition \eqref{equation:linear-separability-1} on $w_1^*, w_2^*, \dots,
w_K^*$. We have
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
\le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}}
\le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2}
\le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
$$
Combining all inequalities, we get
$$
(\gamma M)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
We conclude that $M \le 2(R/\gamma)^2$. Since $M$ is an integer, $M \le \lfloor 2(R/\gamma)^2 \rfloor$.
\end{proof}

@@

It remains to show that the examples we have constructed are linearly separable
with margin $\gamma$. To prove that we demonstrate vectors $w_1, w_2, \dots, w_K$
satisfying conditions \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We define
$$
w_i = \frac{1}{R\sqrt{T}} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} x_t \qquad \qquad \text{for $i=1,2,\dots,K$.}
$$
Let $a_i = |\{ t ~:~ 1 \le t \le T, \ y_t = i \}|$ be the number of occurrences of label $i$.
It is easy to see that
$$
\norm{w_i}^2 = \frac{1}{R^2 T} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} \norm{x_t}^2 = \frac{a_i}{T} \qquad \qquad \text{for $i=1,2,\dots,K$.}
$$
Since $\sum_{i=1}^K a_i = T$, the condition
\eqref{equation:linear-separability-1} holds. To verify condition
\eqref{equation:linear-separability-2} consider any labeled example $(x_t,
y_t)$. Then, by definition of $w_{y_t}$ we have
$$
\ip{w_{y_t}}{x_t}
= \frac{1}{R\sqrt{T}} \sum_{\substack{s : 1 \le s \le T \\ y_s = y_t}} \ip{x_s}{x_t}
= \frac{1}{R\sqrt{T}} \norm{x_t}^2
= \frac{R}{\sqrt{T}}
\ge \frac{R}{R/\gamma}
= \gamma \; .
$$
and for any $i \in \{1,2,\dots,K\} \setminus \{y_t\}$ we have
$\ip{w_i}{x_t} = 0$. Therefore, condition \eqref{equation:linear-separability-2} holds.
\end{proof}

>>>>

\begin{proof}
Let $M = \sum_{t=1}^T \indicator{\widehat y_t \neq y_t}$ be the number of
mistakes the algorithm makes. Since the $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots,
w_K^{(t)})$ changes only if a mistake is made, we can upper bound $\sum_{i=1}^K
\norm{w_i^{(t)}}^2$ in terms of number of mistakes.
If a mistake happens in round $t$ then
\begin{align*}
& \sum_{i=1}^K \norm{w_i^{(t+1)}}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) \\
& \qquad + \norm{w_{y_t}^{(t)} + x_t}^2 + \norm{w_{\widehat y_t}^{(t)} - x_t}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) + \norm{w_{y_t}^{(t)}}^2 + \norm{w_{\widehat y_t}^{(t)}}^2 \\
& \qquad + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& = \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 R^2 \; .
\end{align*}
So each time a mistake happens, $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ increases by at most $2R^2$. Thus,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors satisfying the
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We lower bound $\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$. This quantity changes
only when a mistakes happens. If mistake happens in round $t$, we have
\begin{align*}
& \sum_{i=1}^K \ip{w_i^*}{w_i^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) \\
& \qquad + \ip{w_{y_t}^*}{w_{y_t}^{(t)} + x_t} + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} - x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{y_t}^* - w_{\widehat y_t}^*}{x_t} \\
& \ge  \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma \; .
\end{align*}
Thus, after $M$ mistakes,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma M \; .
$$
We upper bound the left hand side by using Cauchy-Schwartz inequality twice and
the condition \eqref{equation:linear-separability-1} on $w_1^*, w_2^*, \dots,
w_K^*$. We have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
& \le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}} \\
& \le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \\
& \le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
\end{align*}
Combining all inequalities, we get
$$
(\gamma M)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
We conclude that $M \le 2(R/\gamma)^2$. Since $M$ is an integer, $M \le \lfloor 2(R/\gamma)^2 \rfloor$.
\end{proof}

@@

It remains to show that the examples we have constructed are linearly separable
with margin $\gamma$. To prove that we demonstrate vectors $w_1, w_2, \dots, w_K$
satisfying conditions \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We define
$$
w_i = \frac{1}{R\sqrt{T}} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} x_t \qquad \text{for $i=1,2,\dots,K$.}
$$
Let $a_i = |\{ t ~:~ 1 \le t \le T, \ y_t = i \}|$ be the number of occurrences of label $i$.
It is easy to see that
$$
\norm{w_i}^2 = \frac{1}{R^2 T} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} \norm{x_t}^2 = \frac{a_i}{T} \qquad \text{for $i=1,2,\dots,K$.}
$$
Since $\sum_{i=1}^K a_i = T$, the condition
\eqref{equation:linear-separability-1} holds. To verify condition
\eqref{equation:linear-separability-2} consider any labeled example $(x_t,
y_t)$. Then, by definition of $w_{y_t}$ we have
\begin{align*}
\ip{w_{y_t}}{x_t}
& = \frac{1}{R\sqrt{T}} \sum_{\substack{s : 1 \le s \le T \\ y_s = y_t}} \ip{x_s}{x_t} \\
& = \frac{1}{R\sqrt{T}} \norm{x_t}^2 \\
& = \frac{R}{\sqrt{T}} \\
& \ge \frac{R}{R/\gamma} \\
& = \gamma \; .
\end{align*}
and for any $i \in \{1,2,\dots,K\} \setminus \{y_t\}$ we have
$\ip{w_i}{x_t} = 0$. Therefore, condition \eqref{equation:linear-separability-2} holds.
\end{proof}
